\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-10-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.Rmd

Survival analysis and the comparison of event rates has been greatly influenced over the last 50 years by the partial likelihood approach of the Cox proportional hazard model [@cox1972regression]. This approach provides a flexible way of assessing the influence of covariates on the hazard function, without the need to specify a parametric survival model. This flexibility comes at the cost of decoupling the baseline hazard from the effect of the covariates. To recover the whole survival curve---or the cumulative incidence function (CIF)---we then need to separately estimate the baseline hazard [@breslow1972discussion]. This in turn leads to stepwise estimates of the survival function that can be difficult to interpret.

From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group. Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can subsequently be used to estimate the CIF. Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the CIF and the survival function estimates also vary smoothly over time.

With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression. Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk. Their approach handles censored data effortlessly and provides users familiar with generalized linear models a natural way of fitting parametric survival models. Moreover, their framework is very flexible: general functions of time can be estimated (e.g. using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach. And since the unit of analysis is a person moment, time-varying covariates can also easily be included in this framework.

Logistic regression had already been used in the context of discrete-time survival modeling [@cox1972regression]. But in the context of continuous-time survival modeling, using the framework of logistic regression opens the door to an extensive array of powerful modeling tools. Indeed, lasso and elastic-net regression can be used to select variables that are associated with the hazard function. Logistic regression can also be replaced by multinomial regression to analyse data with multiple competing events.

In this article, we present the \CRANpkg{casebase} package [@casebase-package] implemented in R, for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures. In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.

>>>>

The semiparametric Cox model has become the default approach to survival analysis 
even though Cox himself later suggested he would prefer to model the hazard function 
directly. In a 1994 interview with Professor Nancy Reid, Sir David Cox was asked 
how he would model a set of censored survival data, to which he responded: "I 
think I would normally want to tackle problems parametrically ... and if you want 
to do things like predict the outcome for a particular patient, it's much more 
convenient to do that parametrically" [@reid1994conversation]. Indeed, the most 
relevant quantity in a clinical setting is often the 5- or 10-year risk of 
experiencing a certain event given the patient's particular circumstances. 
Unfortunately, the most reported metric from a Cox model is the (potentially 
time-dependent) hazard ratio (HR), which ignores the duration of follow-up or 
is subject to selection bias [@hernan2010hazards]. The covariate adjusted 
survival curve overcomes these limitations, and is arguably a more important 
summary measure to report than the HR. While stepwise survival curves can be 
computed with the Cox model, they require a second step to separately estimate 
the baseline hazard [@breslow1972discussion].

Several authors have since pursued fully parametric approaches that made the 
fitting of smooth hazard functions much more transparent and intuitive through 
generalized linear models. The key feature of these procedures is splitting the 
time axis into discrete intervals. [@whitehead1980fitting] showed the 
equivalence between a Cox model and a Poisson regression with a parameter for 
each event time. [@copenhagen2012needs] provides a nice exposition of this 
equivalence with a real data example and supporting R code for computing 
standard errors. [@arjas1987logistic] and [@efron1988logistic] treated each 
patient-day as a Bernoulli random variable with probability equal to the 
discrete hazard rate. A potential issue with these approaches is that the 
number of time bins need to be chosen by the data analyst. A fine grouping 
of times may result in few (or none) events per interval, which then leads 
to instability in the Newton-Raphson procedure for estimation 
[@kalbfleisch2011statistical]. A coarse grouping could potentially mask 
nonlinear trends in the hazard function.

Rather than discretizing time, Hanley & Miettinen [-@hanley2009fitting] 
selected a discrete set of person-time coordinates (`person-moments') in 
continuous time from all observed follow-up experience constituting the 
study base to obtain the logistic form likelihood expression for the hazard 
function. More specifically, all person-moments when the event of interest 
occurred are selected as the case series, complemented by a randomly sampled 
base series of person-moments serving as controls. This approach allows 
flexible modeling of the hazard function by including time as a covariate 
(e.g. using splines or general additive models). Furthermore, time-dependent 
covariates can be modeled through interactions with time. In short, Hanley & 
Miettinen [-@hanley2009fitting] use the well-understood logistic regression 
for directly modeling the hazard function, but do not require an arbitrary 
choice of splitting time.

In this article, we present the \CRANpkg{casebase} R package [@casebase-package] 
which implements the Hanley & Miettinen [-@hanley2009fitting] approach for 
fitting fully parametric hazard models and covariate adjusted survival curves 
using the familiar interface of the \code{glm} function. Our implementation 
allows for straightforward extensions to other models such as penalized 
regression for variable selection and competing-risk analysis. In addition, 
we provide functions for exploratory data analysis and visualizing the estimated 
quantities such as the hazard function, survival curve and its standard errors. 
The ultimate goal of our package is to make fitting flexible hazards accessible 
to more endusers with the hopes that they will favor reporting absolute risks 
over hazard ratios.

### 2021-10-09 analysis/paper/bhatnagar-turgeon-islam-hanley-saarela.tex

Survival analysis and the comparison of event rates has been greatly
influenced over the last 50 years by the partial likelihood approach of
the Cox proportional hazard model \citep{cox1972regression}. This
approach provides a flexible way of assessing the influence of
covariates on the hazard function, without the need to specify a
parametric survival model. This flexibility comes at the cost of
decoupling the baseline hazard from the effect of the covariates. To
recover the whole survival curve---or the cumulative incidence function
(CIF)---we then need to separately estimate the baseline hazard
\citep{breslow1972discussion}. This in turn leads to stepwise estimates
of the survival function that can be difficult to interpret.

From the perspective of clinicians and their patients, the most relevant
quantity is often the 5- or 10-year risk of experiencing a certain event
given the patient's particular circumstances, and not the hazard ratio
between a treatment and control group. Therefore, to make sound clinical
decisions, it is important to accurately estimate the \emph{full} hazard
function, which can subsequently be used to estimate the CIF. Using a
parametric estimator of the hazard function leads to a smooth function
of time; as a consequence, the CIF and the survival function estimates
also vary smoothly over time.

With the goal of fitting smooth-in-time hazard functions, Hanley \&
Miettinen \citeyearpar{hanley2009fitting} proposed a general framework
for estimating fully parametric hazard models via logistic regression.
Their main idea is simple: comparing person-moments when the event of
interest occurred with moments when patients were at risk. Their
approach handles censored data effortlessly and provides users familiar
with generalized linear models a natural way of fitting parametric
survival models. Moreover, their framework is very flexible: general
functions of time can be estimated (e.g.~using splines or general
additive models), and hence these models retain some of the flexibility
of Cox's partial likelihood approach. And since the unit of analysis is
a person moment, time-varying covariates can also easily be included in
this framework.

Logistic regression had already been used in the context of
discrete-time survival modeling \citep{cox1972regression}. But in the
context of continuous-time survival modeling, using the framework of
logistic regression opens the door to an extensive array of powerful
modeling tools. Indeed, lasso and elastic-net regression can be used to
select variables that are associated with the hazard function. Logistic
regression can also be replaced by multinomial regression to analyse
data with multiple competing events.

In this article, we present the \CRANpkg{casebase} package
\citep{casebase-package} implemented in R, for parametric survival
analysis that combines the ideas of Hanley \& Miettinen
\citeyearpar{hanley2009fitting} into a simple interface. The purpose of
the \pkg{casebase} package is to provide practitioners with an
easy-to-use software tool to compute a patient's risk (or cumulative
incidence) of an event, conditional on a particular patient's covariate
profile. Our package retains the flexibility of case-base sampling and
the familiar interface of the \code{glm} function. It also provides
tools for variable selection, competing-risk analysis, and dealing with
time-dependent exposures. In addition, we provide functions for
exploratory data analysis and visualizing the estimated quantities.

>>>>

The semiparametric Cox model has become the default approach to survival
analysis even though Cox himself later suggested he would prefer to
model the hazard function directly. In a 1994 interview with Professor
Nancy Reid, Sir David Cox was asked how he would model a set of censored
survival data, to which he responded: ``I think I would normally want to
tackle problems parametrically \ldots{} and if you want to do things
like predict the outcome for a particular patient, it's much more
convenient to do that parametrically'' \citep{reid1994conversation}.
Indeed, the most relevant quantity in a clinical setting is often the 5-
or 10-year risk of experiencing a certain event given the patient's
particular circumstances. Unfortunately, the most reported metric from a
Cox model is the (potentially time-dependent) hazard ratio (HR), which
ignores the duration of follow-up or is subject to selection bias
\citep{hernan2010hazards}. The covariate adjusted survival curve
overcomes these limitations, and is arguably a more important summary
measure to report than the HR. While stepwise survival curves can be
computed with the Cox model, they require a second step to separately
estimate the baseline hazard \citep{breslow1972discussion}.

Several authors have since pursued fully parametric approaches that made
the fitting of smooth hazard functions much more transparent and
intuitive through generalized linear models. The key feature of these
procedures is splitting the time axis into discrete intervals.
\citep{whitehead1980fitting} showed the equivalence between a Cox model
and a Poisson regression with a parameter for each event time.
\citep{copenhagen2012needs} provides a nice exposition of this
equivalence with a real data example and supporting R code for computing
standard errors. \citep{arjas1987logistic} and \citep{efron1988logistic}
treated each patient-day as a Bernoulli random variable with probability
equal to the discrete hazard rate. A potential issue with these
approaches is that the number of time bins need to be chosen by the data
analyst. A fine grouping of times may result in few (or none) events per
interval, which then leads to instability in the Newton-Raphson
procedure for estimation \citep{kalbfleisch2011statistical}. A coarse
grouping could potentially mask nonlinear trends in the hazard function.

Rather than discretizing time, Hanley \& Miettinen
\citeyearpar{hanley2009fitting} selected a discrete set of person-time
coordinates (`person-moments') in continuous time from all observed
follow-up experience constituting the study base to obtain the logistic
form likelihood expression for the hazard function. More specifically,
all person-moments when the event of interest occurred are selected as
the case series, complemented by a randomly sampled base series of
person-moments serving as controls. This approach allows flexible
modeling of the hazard function by including time as a covariate
(e.g.~using splines or general additive models). Furthermore,
time-dependent covariates can be modeled through interactions with time.
In short, Hanley \& Miettinen \citeyearpar{hanley2009fitting} use the
well-understood logistic regression for directly modeling the hazard
function, but do not require an arbitrary choice of splitting time.

In this article, we present the \CRANpkg{casebase} R package
\citep{casebase-package} which implements the Hanley \& Miettinen
\citeyearpar{hanley2009fitting} approach for fitting fully parametric
hazard models and covariate adjusted survival curves using the familiar
interface of the \code{glm} function. Our implementation allows for
straightforward extensions to other models such as penalized regression
for variable selection and competing-risk analysis. In addition, we
provide functions for exploratory data analysis and visualizing the
estimated quantities such as the hazard function, survival curve and its
standard errors. The ultimate goal of our package is to make fitting
flexible hazards accessible to more endusers with the hopes that they
will favor reporting absolute risks over hazard ratios.
